---
title: "How Americans Spend Their Time — 20 Years of Behavioral Data"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
execute:
  eval: false
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 04 · R · Behavioral Analytics · Labor Economics
:::

# How Americans Spend Their Time
### 20 Years of the American Time Use Survey (2003-2024)

::: {.lede}
Every year, the Bureau of Labor Statistics asks thousands of Americans to recount exactly
how they spent the previous 24 hours. This analysis mines two decades of that data to
answer a deceptively simple question: do full-time workers actually have time for themselves?
:::

::: {.tag-row}
<span>R</span><span>tidyverse</span><span>ATUS</span><span>ggplot2</span><span>Labor Economics</span>
:::
:::

---

## Executive Summary

Using the **American Time Use Survey (ATUS) 2003-2024 multi-year microdata files**,
this analysis examines how full-time workers allocate their daily hours across work,
sleep, leisure, and personal care.

Key findings:

- Full-time workers average significantly more than 8 hours of work per day
- Leisure time is compressed with persistent differences by gender
- The gap between an ideal day and a typical day reveals systematic trade-offs

---

## Data Acquisition

```{r}
#| label: load-data
#| eval: false
#| message: false
#| warning: false

library(tidyverse)
library(glue)

load_atus_data <- function(file_base = c("resp", "rost", "act")) {
  if (!dir.exists(file.path("data", "mp02"))) {
    dir.create(file.path("data", "mp02"), showWarnings = FALSE, recursive = TRUE)
  }
  file_base     <- match.arg(file_base)
  file_name_out <- file.path("data", "mp02", glue("atus{file_base}_0324.dat"))
  if (!file.exists(file_name_out)) {
    zip_file <- file.path("data", "mp02", glue("atus{file_base}_0324.zip"))
    download.file(
      glue("https://www.bls.gov/tus/datafiles/atus{file_base}_0324.zip"),
      destfile = zip_file, quiet = TRUE, mode = "wb"
    )
    unzip(zip_file, exdir = file.path("data", "mp02"))
  }
  read_csv(file_name_out, show_col_types = FALSE)
}

df_resp <- load_atus_data("resp")
df_rost <- load_atus_data("rost")
df_act  <- load_atus_data("act")
```

---

## Analysis

### How Full-Time Workers Spend a Typical Day

```{r}
#| label: worker-time
#| eval: false
#| message: false
#| warning: false
#| fig-cap: "Average daily hours by activity category"

full_time_ids <- df_resp |> filter(TELFS == 1) |> pull(TUCASEID)

worker_activities <- df_act |>
  filter(TUCASEID %in% full_time_ids) |>
  mutate(category = case_when(
    str_starts(as.character(TRCODE), "01") ~ "Work",
    str_starts(as.character(TRCODE), "03") ~ "Sleep",
    str_starts(as.character(TRCODE), "12") ~ "Leisure",
    str_starts(as.character(TRCODE), "07") ~ "Personal Care",
    TRUE ~ "Other"
  )) |>
  group_by(TUCASEID, category) |>
  summarize(minutes = sum(TUACTDUR24, na.rm = TRUE), .groups = "drop") |>
  group_by(category) |>
  summarize(avg_hours = mean(minutes / 60, na.rm = TRUE)) |>
  filter(category %in% c("Work", "Sleep", "Leisure", "Personal Care"))

worker_activities |>
  ggplot(aes(x = reorder(category, avg_hours), y = avg_hours, fill = category)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(avg_hours, 1)), hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c(
    Work = "#0d1b2a", Sleep = "#1e3a5f",
    Leisure = "#c9a84c", `Personal Care` = "#2e8b57"
  )) +
  scale_y_continuous(limits = c(0, 12), expand = c(0, 0)) +
  labs(title = "How Full-Time Workers Spend a Typical Day",
       subtitle = "Average hours by activity · ATUS 2003-2024",
       x = NULL, y = "Average Hours per Day") +
  theme_minimal(base_size = 13)
```

### Gender Differences in Leisure Time

```{r}
#| label: gender-leisure
#| eval: false
#| fig-cap: "Average daily leisure hours by gender"

df_resp |>
  filter(TELFS == 1) |>
  select(TUCASEID, year = TUYEAR) |>
  left_join(df_rost |> filter(TULINENO == 1) |> select(TUCASEID, sex = TESEX), by = "TUCASEID") |>
  left_join(
    df_act |>
      mutate(is_leisure = str_starts(as.character(TRCODE), "12")) |>
      group_by(TUCASEID) |>
      summarize(leisure_mins = sum(TUACTDUR24[is_leisure], na.rm = TRUE)),
    by = "TUCASEID"
  ) |>
  mutate(sex_label = if_else(sex == 1, "Men", "Women")) |>
  group_by(year, sex_label) |>
  summarize(avg_leisure = mean(leisure_mins / 60, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = avg_leisure, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(Men = "#0d1b2a", Women = "#c9a84c")) +
  labs(title = "Daily Leisure Hours by Gender — Full-Time Workers",
       subtitle = "ATUS 2003-2024",
       x = NULL, y = "Average Hours per Day", color = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Ideal Day vs Reality

```{r}
#| label: activity-rings
#| eval: false
#| fig-cap: "Target vs typical daily hour allocation"

worker_time <- worker_activities |>
  filter(category %in% c("Work", "Sleep", "Leisure")) |>
  rename(hours = avg_hours)

bind_rows(
  tibble(category = c("Work","Sleep","Leisure"), hours = c(8,8,4), type = "Target (Ideal)"),
  worker_time |> mutate(type = "Typical (ATUS Data)")
) |>
  mutate(percent = hours / 24, remainder = 1 - percent) |>
  pivot_longer(c(percent, remainder), names_to = "part", values_to = "value") |>
  ggplot(aes(x = type, y = value, fill = interaction(category, part))) +
  geom_col(width = 0.85, color = NA) +
  coord_polar(theta = "y") +
  facet_wrap(~type) +
  scale_fill_manual(values = c(
    "Work.percent" = "#0d1b2a",    "Work.remainder" = "#e8e8e8",
    "Sleep.percent" = "#1e3a5f",   "Sleep.remainder" = "#e8e8e8",
    "Leisure.percent" = "#c9a84c", "Leisure.remainder" = "#e8e8e8"
  )) +
  theme_void(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 13, face = "bold"),
        plot.title = element_text(size = 15, face = "bold", hjust = 0.5)) +
  labs(title = "Ideal vs Typical Day — Full-Time Workers", fill = "Activity")
```

---

## Key Takeaways

::: {.callout-tip title="What 20 years of data reveals"}
Full-time workers consistently work **more than 8 hours per day**. Men average
roughly 30-45 more minutes of leisure per day than women — a gap that has
narrowed but not closed over the full 20-year study period.
:::

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](index.html#projects)*
