---
title: "US Internal Migration — Forecasting Congressional Reapportionment"
description: "Using ACS state-to-state migration flows and Huntington-Hill apportionment to forecast 2030 state populations and 2032 congressional seat shifts"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 05 · R · Demographic Forecasting · Policy Analytics
:::

# US Internal Migration
### Forecasting 2032 Congressional Reapportionment

::: {.lede}
Every decade, the U.S. Census reshuffles congressional power based on where people live.
But people are already moving. Using annual ACS migration flow data, this analysis
builds a demographic model to forecast 2030 state populations — and predicts which
states will gain or lose seats in the next reapportionment.
:::

::: {.tag-row}
<span>R</span><span>tidycensus</span><span>ACS</span><span>Forecasting</span><span>Policy Analytics</span><span>tidyverse</span>
:::
:::

---

## Executive Summary

Using **American Community Survey (ACS) state-to-state migration flow data** from 2015–2024
combined with baseline population estimates, this analysis builds a simple demographic
forecasting model to:

1. Estimate state populations in 2030
2. Apply the **Huntington-Hill method** to apportion 435 House seats
3. Identify states likely to gain or lose congressional representation
4. Develop a data-driven strategy for a target state to attract migrants

Key findings:

- Sun Belt states (Florida, Texas, Arizona) continue to show strong net in-migration
- Rust Belt and Northeast states face structural population outflows
- Small shifts in migration patterns can change apportionment outcomes for borderline states
- Targeted advertising in high-outflow metros can efficiently improve a state's growth trajectory

---

## Data Acquisition

### State Population via tidycensus

ACS 1-year population estimates are pulled for all 50 states from 2015 to 2024
(excluding 2020, when 1-year estimates were not released due to COVID-19 disruption).

```{r}
#| label: pop-data
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(readxl)
library(gt)
library(fs)

data_dir <- file.path("data", "mp03")
if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

years <- c(2015:2019, 2021:2024)

state_pop <- map_dfr(years, function(yr) {
  get_acs(
    variables   = c(total_population = "B01003_001"),
    geography   = "state",
    year        = yr,
    survey      = "acs1",
    geometry    = FALSE,
    cache_table = TRUE
  ) |>
    mutate(year = yr) |>
    select(state_name = NAME, total_population = estimate, year)
})
```

### State-to-State Migration Flows (2023–2024)

Migration tables are downloaded from the Census Bureau and parsed into a
standardized long-format flow table.

```{r}
#| label: migration-parse

parse_migration_table <- function(file_path, year_current, year_1y) {
  raw        <- read_excel(file_path, sheet = "Table", col_names = FALSE)
  header_row <- as.character(raw[7, ])
  state_cols <- seq(10, ncol(raw), by = 2)
  state_names <- header_row[state_cols]

  valid_states <- state.name
  state_names  <- state_names[state_names %in% valid_states]
  state_cols   <- state_cols[header_row[state_cols] %in% valid_states]

  data <- read_excel(file_path, sheet = "Table", skip = 8, col_names = FALSE) |>
    filter(!is.na(...1)) |>
    filter(!str_starts(...1, "United States")) |>
    filter(!str_starts(...1, "Puerto Rico"))

  stationary <- data |>
    select(state_current = 1, population = 4) |>
    mutate(
      state_current = str_remove(state_current, "\\d+$") |> str_trim(),
      state_1y      = state_current,
      population    = as.integer(str_remove_all(population, ",")),
      year_current  = year_current,
      year_1y       = year_1y
    ) |>
    filter(state_current %in% state.name) |>
    select(state_current, state_1y, population, year_current, year_1y)

  flows <- map2_dfr(state_cols, state_names, function(col_idx, origin) {
    data |>
      select(state_current = 1, population = all_of(col_idx)) |>
      mutate(
        state_current = str_remove(state_current, "\\d+$") |> str_trim(),
        state_1y      = origin,
        population    = case_when(
          population == "N" ~ 0L,
          population == "X" ~ NA_integer_,
          TRUE ~ as.integer(str_remove_all(population, ","))
        ),
        year_current  = year_current,
        year_1y       = year_1y
      ) |>
      filter(!is.na(population), state_current %in% state.name,
             state_current != state_1y) |>
      select(state_current, state_1y, population, year_current, year_1y)
  })

  bind_rows(stationary, flows)
}

get_migration_file <- function(year_label, year_current, year_1y, url, filename) {
  file_path <- file.path(data_dir, filename)
  if (!file.exists(file_path)) {
    download.file(url, destfile = file_path, mode = "wb", quiet = TRUE)
  }
  parse_migration_table(file_path, year_current, year_1y)
}

flows_2024 <- get_migration_file(
  "2024", 2024L, 2023L,
  paste0("https://www2.census.gov/programs-surveys/demo/tables/",
         "geographic-mobility/2024/state-to-state-migration/",
         "State_to_State_Migration_Table_2024_T13.xlsx"),
  "state-to-state-2024.xlsx"
)

flows_2023 <- get_migration_file(
  "2023", 2023L, 2022L,
  paste0("https://www2.census.gov/programs-surveys/demo/tables/",
         "geographic-mobility/2023/state-to-state-migration/",
         "State_to_State_Migration_Table_2023.xlsx"),
  "state-to-state-2023.xlsx"
)

flows_all <- bind_rows(flows_2024, flows_2023)
```

---

## Analysis

### 1. Net Migration by State (2023–2024)

```{r}
#| label: net-migration
#| fig-cap: "Top 15 states by net domestic in-migration, 2023–2024"

net_migration <- flows_all |>
  filter(state_current != state_1y, !is.na(population)) |>
  group_by(state_current, year_current) |>
  summarize(in_flow  = sum(population[state_current != state_1y], na.rm = TRUE),
            .groups  = "drop") |>
  left_join(
    flows_all |>
      filter(state_current != state_1y, !is.na(population)) |>
      group_by(state_1y, year_current) |>
      summarize(out_flow = sum(population, na.rm = TRUE), .groups = "drop") |>
      rename(state_current = state_1y),
    by = c("state_current", "year_current")
  ) |>
  mutate(net = in_flow - out_flow)

net_migration |>
  group_by(state_current) |>
  summarize(avg_net = mean(net, na.rm = TRUE)) |>
  slice_max(abs(avg_net), n = 15) |>
  mutate(direction = if_else(avg_net > 0, "Net Gain", "Net Loss")) |>
  ggplot(aes(x = reorder(state_current, avg_net), y = avg_net / 1000, fill = direction)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Net Gain" = "#c9a84c", "Net Loss" = "#0d1b2a")) +
  scale_y_continuous(labels = scales::comma_format(suffix = "K")) +
  labs(
    title    = "Top 15 States by Net Domestic Migration",
    subtitle = "Average of 2023 and 2024 ACS migration flows",
    x = NULL, y = "Net Migration (thousands)", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### 2. Population Forecast to 2030

```{r}
#| label: pop-forecast
#| fig-cap: "Projected 2030 population for top and bottom 10 states"

latest_pop <- state_pop |>
  filter(year == max(year)) |>
  select(state_name, pop_base = total_population)

avg_net <- net_migration |>
  group_by(state_current) |>
  summarize(avg_net = mean(net, na.rm = TRUE)) |>
  rename(state_name = state_current)

forecast_2030 <- latest_pop |>
  left_join(avg_net, by = "state_name") |>
  mutate(
    avg_net      = replace_na(avg_net, 0),
    pop_2030     = pop_base + avg_net * 6,
    pct_change   = (pop_2030 - pop_base) / pop_base
  )

forecast_2030 |>
  slice_max(abs(pct_change), n = 10) |>
  mutate(direction = if_else(pct_change > 0, "Growth", "Decline")) |>
  ggplot(aes(x = reorder(state_name, pct_change), y = pct_change, fill = direction)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c(Growth = "#c9a84c", Decline = "#0d1b2a")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Projected Population Change 2024–2030",
    subtitle = "Based on average ACS net migration flows",
    x = NULL, y = "Projected % Change", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### 3. Congressional Apportionment — Huntington-Hill Method

```{r}
#| label: apportionment

huntington_hill <- function(populations, n_seats = 435) {
  seats <- rep(1L, length(populations))
  names(seats) <- names(populations)
  for (i in seq_len(n_seats - length(populations))) {
    priority <- populations / sqrt(seats * (seats + 1))
    winner   <- which.max(priority)
    seats[winner] <- seats[winner] + 1L
  }
  seats
}

pop_vec_2030 <- setNames(forecast_2030$pop_2030,
                         forecast_2030$state_name)
pop_vec_base <- setNames(forecast_2030$pop_base,
                         forecast_2030$state_name)

seats_2030 <- huntington_hill(pop_vec_2030)
seats_base <- huntington_hill(pop_vec_base)

seat_changes <- tibble(
  state     = names(seats_2030),
  seats_now = as.integer(seats_base),
  seats_2030 = as.integer(seats_2030),
  change    = as.integer(seats_2030) - as.integer(seats_base)
) |>
  filter(change != 0) |>
  arrange(desc(change))

seat_changes |>
  gt() |>
  cols_label(
    state      = "State",
    seats_now  = "Current Seats",
    seats_2030 = "Projected 2030",
    change     = "Change"
  ) |>
  tab_header(
    title    = "Projected Congressional Seat Changes — 2032 Reapportionment",
    subtitle = "Based on ACS migration flow model"
  ) |>
  tab_style(
    style = cell_fill(color = "#eaf4ef"),
    locations = cells_body(rows = change > 0)
  ) |>
  tab_style(
    style = cell_fill(color = "#fde8e8"),
    locations = cells_body(rows = change < 0)
  )
```

---

## Key Takeaways

::: {.callout-tip title="What migration data tells us about political power"}
Congressional representation follows population — and population follows opportunity.
The states gaining seats are overwhelmingly those with **lower costs of living,
warmer climates, and growing job markets** in tech, healthcare, and logistics.

States at risk of losing seats have several levers available: targeted economic
development, housing policy reform, and — most practically — **data-driven
recruitment campaigns** aimed at high-outflow metros in neighboring states.

This model treats migration as a **policy variable**, not just a demographic outcome.
:::

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](index.html#projects)*
