---
title: "Campus Diversity After SFFA — A National Enrollment Analysis"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 03 · R · Policy Analytics · Education Data
:::

# Campus Diversity After SFFA
### A National Enrollment Analysis (2010–2024)

::: {.lede}
In June 2023, the Supreme Court ruling in Students for Fair Admissions v. Harvard
ended race-conscious admissions at U.S. universities. This analysis uses 15 years of
federal enrollment data to measure what actually changed and what was already in motion
long before the ruling.
:::

::: {.tag-row}
<span>R</span><span>tidyverse</span><span>IPEDS</span><span>gt</span><span>Policy Analysis</span>
:::
:::

---

## Executive Summary

Using enrollment microdata from the **Integrated Postsecondary Education Data System (IPEDS)**
across 3,000+ institutions from 2010 to 2024, this analysis examines whether the SFFA ruling
produced a measurable shift in campus racial and ethnic composition.

Key findings:

- Many diversity trends predate 2023 and reflect regional population shifts
- Baruch College shows a distinct demographic trajectory vs national averages
- CUNY institutions enroll a significant share of New York first-year undergraduates

---

## Data Acquisition

```{r}
#| label: acquire-data
#| message: false
#| warning: false

library(tidyverse)
library(glue)
library(gt)

acquire_ipeds_data <- function(start_year = 2010, end_year = 2024) {
  data_dir <- file.path("data", "mp01")
  if (!dir.exists(data_dir)) dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
  YEARS <- seq(start_year, end_year)

  EFA_ALL <- map(YEARS, function(yy) {
    ef_url <- if (yy <= 2022) {
      glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
    } else {
      glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
    }
    ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
    if (!file.exists(ef_file)) download.file(ef_url, destfile = ef_file, quiet = TRUE, mode = "wb")
    read_csv(ef_file, show_col_types = FALSE) |>
      mutate(
        year = yy,
        enrollment_m_asia = EFASIAM, enrollment_f_asia = EFASIAW,
        enrollment_m_bkaa = EFBKAAM, enrollment_f_bkaa = EFBKAAW,
        enrollment_m_hisp = EFHISPM, enrollment_f_hisp = EFHISPW,
        enrollment_m_whit = EFWHITM, enrollment_f_whit = EFWHITW,
        enrollment_m_aian = EFAIANM, enrollment_f_aian = EFAIANW,
        enrollment_m_nhpi = EFNHPIM, enrollment_f_nhpi = EFNHPIW,
        enrollment_m_2mor = EF2MORM, enrollment_f_2mor = EF2MORW,
        enrollment_m_unkn = EFUNKNM, enrollment_f_unkn = EFUNKNW,
        enrollment_m_nral = EFNRALM, enrollment_f_nral = EFNRALW
      ) |>
      filter((EFALEVEL %in% c(2, 12)) | (LINE %in% c(1, 15))) |>
      mutate(level = case_when(
        EFALEVEL == 2       ~ "all undergrad",
        EFALEVEL == 12      ~ "all graduate",
        LINE %in% c(1, 15) ~ "first year undergrad"
      )) |>
      select(institution_id = UNITID, year, level, starts_with("enrollment_")) |>
      group_by(institution_id, year, level) |>
      summarize(across(starts_with("enrollment_"), sum), .groups = "drop")
  }) |> bind_rows()

  DESC_ALL <- map(YEARS, function(yy) {
    hd_url <- if (yy <= 2022) {
      glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
    } else {
      glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
    }
    hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
    if (!file.exists(hd_file)) download.file(hd_url, destfile = hd_file, quiet = TRUE, mode = "wb")
    suppressWarnings(
      read_csv(hd_file, show_col_types = FALSE,
               locale = locale(encoding = if_else(yy == 2024, "utf-8", "windows-1252"))) |>
        mutate(year = yy) |>
        select(institution_id = UNITID, institution_name = INSTNM, state = STABBR, year)
    )
  }) |> bind_rows()

  inner_join(EFA_ALL, DESC_ALL, join_by(institution_id, year))
}

IPEDS <- acquire_ipeds_data() |>
  mutate(
    total_students = rowSums(across(starts_with("enrollment_")), na.rm = TRUE),
    total_female   = rowSums(across(matches("enrollment_f_")), na.rm = TRUE),
    asian_students = enrollment_m_asia + enrollment_f_asia,
    black_students = enrollment_m_bkaa + enrollment_f_bkaa,
    hisp_students  = enrollment_m_hisp + enrollment_f_hisp,
    white_students = enrollment_m_whit + enrollment_f_whit
  )

distinct_institutions <- IPEDS |> distinct(institution_id) |> nrow()
```

The dataset covers **`r scales::comma(distinct_institutions)` institutions** across all 50 states.

---

## Analysis

### National Enrollment Trends by Race/Ethnicity

```{r}
#| label: national-trends
#| fig-cap: "Undergraduate enrollment share by race/ethnicity, 2010-2024"

IPEDS |>
  filter(level == "all undergrad") |>
  group_by(year) |>
  summarize(
    Asian    = sum(asian_students, na.rm = TRUE),
    Black    = sum(black_students, na.rm = TRUE),
    Hispanic = sum(hisp_students,  na.rm = TRUE),
    White    = sum(white_students, na.rm = TRUE),
    Total    = sum(total_students, na.rm = TRUE)
  ) |>
  mutate(across(Asian:White, ~ .x / Total)) |>
  select(-Total) |>
  pivot_longer(-year, names_to = "group", values_to = "share") |>
  ggplot(aes(x = year, y = share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = c(
    Asian = "#c9a84c", Black = "#1e3a5f",
    Hispanic = "#2e8b57", White = "#5a6a7e"
  )) +
  labs(title = "National Undergraduate Enrollment Share by Race/Ethnicity",
       subtitle = "IPEDS Fall Enrollment 2010-2024",
       x = NULL, y = "Share of Total Enrollment", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### First-Year Enrollment at CUNY (2024)

```{r}
#| label: cuny-enrollment
IPEDS |>
  filter(year == 2024, level == "first year undergrad",
         str_detect(institution_name, "CUNY")) |>
  group_by(institution_name) |>
  summarize(first_year = sum(total_students)) |>
  mutate(pct_total = first_year / sum(first_year)) |>
  arrange(desc(first_year)) |>
  gt() |>
  cols_label(institution_name = "Institution",
             first_year = "First-Year Students",
             pct_total = "Share of CUNY Total") |>
  tab_header(title = "First-Year Undergraduate Enrollment at CUNY (2024)") |>
  fmt_percent(pct_total, decimals = 1) |>
  fmt_number(first_year, decimals = 0)
```

### Baruch College Enrollment Over Time

```{r}
#| label: baruch-trend
IPEDS |>
  filter(str_detect(institution_name, "Baruch"), level == "all undergrad") |>
  group_by(year) |>
  summarize(total_undergrad = sum(total_students)) |>
  arrange(year) |>
  mutate(yoy_change = (total_undergrad - lag(total_undergrad)) / lag(total_undergrad)) |>
  gt() |>
  cols_label(year = "Year", total_undergrad = "Enrollment", yoy_change = "YoY Change") |>
  tab_header(title = "Baruch College Undergraduate Enrollment (2010-2024)") |>
  fmt_percent(yoy_change, decimals = 1) |>
  fmt_number(total_undergrad, decimals = 0)
```

---

## Key Takeaways

::: {.callout-tip title="What the data shows"}
Demographic shifts in U.S. higher education are **long-term structural trends**.
Hispanic and multiracial enrollment has grown steadily since 2010 driven by
changing U.S. population demographics, not admissions policy alone.
:::

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](index.html#projects)*
