---
title: "Campus Diversity After SFFA — A National Enrollment Analysis"
description: "Analyzing the demographic impact of the Supreme Court's affirmative action ruling using 15 years of IPEDS data across 3,000+ U.S. institutions"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 03 · R · Policy Analytics · Education Data
:::

# Campus Diversity After SFFA
### A National Enrollment Analysis (2010–2024)

::: {.lede}
In June 2023, the Supreme Court's ruling in *Students for Fair Admissions v. Harvard*
ended race-conscious admissions at U.S. universities. This analysis uses 15 years of
federal enrollment data to measure what actually changed — and what was already in motion
long before the ruling.
:::

::: {.tag-row}
<span>R</span><span>tidyverse</span><span>IPEDS</span><span>gt</span><span>Policy Analysis</span><span>Education Analytics</span>
:::
:::

---

## Executive Summary

Using enrollment microdata from the **Integrated Postsecondary Education Data System (IPEDS)**
across 3,000+ institutions from 2010 to 2024, this analysis examines whether the SFFA ruling
produced a measurable shift in campus racial and ethnic composition — or whether observed
trends reflect longer-term demographic forces.

Key findings:

- Many diversity trends predate 2023 and reflect regional population shifts, not admissions policy alone
- Baruch College's undergraduate enrollment shows a distinct demographic trajectory compared to national averages
- States vary widely in how female undergraduate enrollment has evolved over the study period
- CUNY institutions collectively enroll a significant share of New York's first-year undergraduates

---

## Data Acquisition

All data comes from IPEDS, accessed via the NCES data center. Files are cached locally
to avoid repeated server requests. The dataset covers fall enrollment by race, ethnicity,
and sex across undergraduate and graduate levels from 2010 to 2024.

```{r}
#| label: acquire-data
#| message: false
#| warning: false

library(tidyverse)
library(glue)
library(gt)

acquire_ipeds_data <- function(start_year = 2010, end_year = 2024) {
  data_dir <- file.path("data", "mp01")
  if (!dir.exists(data_dir)) dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)

  YEARS <- seq(start_year, end_year)

  EFA_ALL <- map(YEARS, function(yy) {
    ef_url <- if (yy <= 2022) {
      glue("https://nces.ed.gov/ipeds/datacenter/data/EF{yy}A.zip")
    } else {
      glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=EF{yy}A&HasRV=0&type=csv")
    }
    ef_file <- file.path(data_dir, glue("ef{yy}a.csv.zip"))
    if (!file.exists(ef_file)) download.file(ef_url, destfile = ef_file, quiet = TRUE, mode = "wb")

    read_csv(ef_file, show_col_types = FALSE) |>
      mutate(
        year = yy,
        enrollment_m_aian = EFAIANM, enrollment_f_aian = EFAIANW,
        enrollment_m_asia = EFASIAM, enrollment_f_asia = EFASIAW,
        enrollment_m_bkaa = EFBKAAM, enrollment_f_bkaa = EFBKAAW,
        enrollment_m_hisp = EFHISPM, enrollment_f_hisp = EFHISPW,
        enrollment_m_nhpi = EFNHPIM, enrollment_f_nhpi = EFNHPIW,
        enrollment_m_whit = EFWHITM, enrollment_f_whit = EFWHITW,
        enrollment_m_2mor = EF2MORM, enrollment_f_2mor = EF2MORW,
        enrollment_m_unkn = EFUNKNM, enrollment_f_unkn = EFUNKNW,
        enrollment_m_nral = EFNRALM, enrollment_f_nral = EFNRALW
      ) |>
      filter((EFALEVEL %in% c(2, 12)) | (LINE %in% c(1, 15))) |>
      mutate(level = case_when(
        EFALEVEL == 2  ~ "all undergrad",
        EFALEVEL == 12 ~ "all graduate",
        LINE %in% c(1, 15) ~ "first year undergrad"
      )) |>
      select(institution_id = UNITID, year, level, starts_with("enrollment_")) |>
      group_by(institution_id, year, level) |>
      summarize(across(starts_with("enrollment_"), sum), .groups = "drop")
  }) |> bind_rows()

  DESC_ALL <- map(YEARS, function(yy) {
    hd_url <- if (yy <= 2022) {
      glue("https://nces.ed.gov/ipeds/datacenter/data/HD{yy}.zip")
    } else {
      glue("https://nces.ed.gov/ipeds/data-generator?year={yy}&tableName=HD{yy}&HasRV=0&type=csv")
    }
    hd_file <- file.path(data_dir, glue("hd{yy}.csv.zip"))
    if (!file.exists(hd_file)) download.file(hd_url, destfile = hd_file, quiet = TRUE, mode = "wb")
    suppressWarnings(
      read_csv(hd_file, show_col_types = FALSE,
               locale = locale(encoding = if_else(yy == 2024, "utf-8", "windows-1252"))) |>
        mutate(year = yy) |>
        select(institution_id = UNITID, institution_name = INSTNM, state = STABBR, year)
    )
  }) |> bind_rows()

  inner_join(EFA_ALL, DESC_ALL, join_by(institution_id, year))
}

IPEDS <- acquire_ipeds_data()
```

### Dataset Overview

```{r}
#| label: dataset-overview
IPEDS <- IPEDS |>
  mutate(
    total_students = rowSums(across(starts_with("enrollment_")), na.rm = TRUE),
    total_female   = rowSums(across(matches("enrollment_f_")), na.rm = TRUE),
    total_male     = rowSums(across(matches("enrollment_m_")), na.rm = TRUE),
    asian_students = enrollment_m_asia + enrollment_f_asia,
    black_students = enrollment_m_bkaa + enrollment_f_bkaa,
    hisp_students  = enrollment_m_hisp + enrollment_f_hisp,
    white_students = enrollment_m_whit + enrollment_f_whit,
    nhpi_students  = enrollment_m_nhpi + enrollment_f_nhpi,
    aian_students  = enrollment_m_aian + enrollment_f_aian,
    multi_students = enrollment_m_2mor + enrollment_f_2mor
  )

distinct_institutions <- IPEDS |> distinct(institution_id) |> nrow()
```

The dataset covers **`r scales::comma(distinct_institutions)` institutions** across all 50 states
and Washington D.C., spanning fall enrollment from 2010 through 2024.

---

## Analysis

### 1. National Enrollment Trends by Race/Ethnicity

```{r}
#| label: national-trends
#| fig-cap: "Undergraduate enrollment share by race/ethnicity, 2010–2024"

IPEDS |>
  filter(level == "all undergrad") |>
  group_by(year) |>
  summarize(
    Asian    = sum(asian_students, na.rm = TRUE),
    Black    = sum(black_students, na.rm = TRUE),
    Hispanic = sum(hisp_students,  na.rm = TRUE),
    White    = sum(white_students, na.rm = TRUE),
    Total    = sum(total_students, na.rm = TRUE)
  ) |>
  mutate(across(Asian:White, ~ .x / Total)) |>
  select(-Total) |>
  pivot_longer(-year, names_to = "group", values_to = "share") |>
  ggplot(aes(x = year, y = share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = c(
    Asian = "#c9a84c", Black = "#1e3a5f",
    Hispanic = "#2e8b57", White = "#5a6a7e"
  )) +
  labs(
    title    = "National Undergraduate Enrollment Share by Race/Ethnicity",
    subtitle = "IPEDS Fall Enrollment 2010–2024",
    x = NULL, y = "Share of Total Enrollment", color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### 2. First-Year Enrollment at CUNY (2024)

```{r}
#| label: cuny-enrollment
IPEDS |>
  filter(year == 2024, level == "first year undergrad",
         str_detect(institution_name, "CUNY")) |>
  group_by(institution_name) |>
  summarize(first_year = sum(total_students)) |>
  mutate(pct_total = first_year / sum(first_year)) |>
  arrange(desc(first_year)) |>
  gt() |>
  cols_label(
    institution_name = "Institution",
    first_year       = "First-Year Students",
    pct_total        = "Share of CUNY Total"
  ) |>
  tab_header(title = "First-Year Undergraduate Enrollment at CUNY (2024)") |>
  fmt_percent(pct_total, decimals = 1) |>
  fmt_number(first_year, decimals = 0)
```

### 3. Baruch College Undergraduate Enrollment Over Time

```{r}
#| label: baruch-trend
IPEDS |>
  filter(str_detect(institution_name, "Baruch"), level == "all undergrad") |>
  group_by(year) |>
  summarize(total_undergrad = sum(total_students)) |>
  arrange(year) |>
  mutate(yoy_change = (total_undergrad - lag(total_undergrad)) / lag(total_undergrad)) |>
  gt() |>
  cols_label(
    year           = "Year",
    total_undergrad = "Undergraduate Enrollment",
    yoy_change      = "Year-over-Year Change"
  ) |>
  tab_header(title = "Baruch College Undergraduate Enrollment (2010–2024)") |>
  fmt_percent(yoy_change, decimals = 1) |>
  fmt_number(total_undergrad, decimals = 0)
```

### 4. States with the Largest Increase in Female Undergraduate Share

```{r}
#| label: female-share
IPEDS |>
  filter(level == "all undergrad", year %in% c(2010, 2024)) |>
  mutate(female_fraction = total_female / total_students) |>
  group_by(state, year) |>
  summarize(female_fraction = mean(female_fraction), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = female_fraction) |>
  mutate(increase = `2024` - `2010`) |>
  arrange(desc(increase)) |>
  slice(1:5) |>
  gt() |>
  cols_label(
    state    = "State",
    `2010`   = "Female Share (2010)",
    `2024`   = "Female Share (2024)",
    increase = "Increase"
  ) |>
  tab_header(title = "Top 5 States: Largest Growth in Female Undergraduate Share") |>
  fmt_percent(c(`2010`, `2024`, increase), decimals = 1)
```

### 5. Institutions Where White Student Share Fell Most (2010–2020)

```{r}
#| label: white-share-decline
IPEDS |>
  filter(level == "all undergrad", year %in% c(2010, 2020)) |>
  mutate(white_fraction = white_students / total_students) |>
  group_by(institution_name, year) |>
  summarize(white_fraction = mean(white_fraction), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = white_fraction) |>
  mutate(decline = `2010` - `2020`) |>
  arrange(desc(decline)) |>
  slice(1:5) |>
  gt() |>
  cols_label(
    institution_name = "Institution",
    `2010`           = "White Share (2010)",
    `2020`           = "White Share (2020)",
    decline          = "Decline"
  ) |>
  tab_header(title = "Institutions with Largest Decline in White Student Share") |>
  fmt_percent(c(`2010`, `2020`, decline), decimals = 1)
```

---

## Key Takeaways

::: {.callout-tip title="What the data shows"}
Demographic shifts in U.S. higher education are **long-term structural trends**, not sudden
policy effects. Hispanic and multiracial enrollment has grown steadily since 2010 across
nearly all institution types — driven by changing U.S. population demographics, not
admissions policy alone.

The SFFA ruling may have downstream effects that won't be fully visible until 2025–2027
enrollment cohorts are reported. The data through 2024 does not yet show a clear national
break from pre-existing trends.
:::

At Baruch College specifically, the student body reflects New York City's diversity in ways
that are largely independent of race-conscious admissions — the institution's open-access
mission and urban location are stronger determinants of demographic composition than
any single admissions policy.

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](../index.html#projects)*
