---
title: "How Americans Spend Their Time — 20 Years of Behavioral Data"
description: "Analyzing 20 years of ATUS microdata to uncover how full-time workers balance work, sleep, and leisure — and what the data says about modern work-life balance"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 04 · R · Behavioral Analytics · Labor Economics
:::

# How Americans Spend Their Time
### 20 Years of the American Time Use Survey (2003–2024)

::: {.lede}
Every year, the Bureau of Labor Statistics asks thousands of Americans to recount exactly
how they spent the previous 24 hours. This analysis mines two decades of that data to
answer a deceptively simple question: do full-time workers actually have time for themselves?
:::

::: {.tag-row}
<span>R</span><span>tidyverse</span><span>ATUS</span><span>ggplot2</span><span>Behavioral Analytics</span><span>Labor Economics</span>
:::
:::

---

## Executive Summary

Using the **American Time Use Survey (ATUS) 2003–2024 multi-year microdata files**,
this analysis examines how full-time workers allocate their daily hours across work,
sleep, leisure, and personal care — and how those patterns have shifted over two decades.

Key findings:

- Full-time workers average significantly more than 8 hours of work per day when commuting and work-related activities are included
- Leisure time for full-time workers has remained compressed, with meaningful differences by gender
- Sleep averages fall below recommended levels for a substantial share of the working population
- The gap between an "ideal" day and a "typical" day reveals systematic trade-offs in how Americans manage time

---

## Data Acquisition

The ATUS multi-year microdata files are downloaded from the BLS ATUS extract system.
Three tables are used: respondent-level data (`resp`), roster demographics (`rost`),
and activity-level records (`act`). Files are cached locally.

```{r}
#| label: load-data
#| message: false
#| warning: false

library(tidyverse)
library(glue)
library(readxl)

load_atus_data <- function(file_base = c("resp", "rost", "act")) {
  if (!dir.exists(file.path("data", "mp02"))) {
    dir.create(file.path("data", "mp02"), showWarnings = FALSE, recursive = TRUE)
  }
  file_base     <- match.arg(file_base)
  file_name_out <- file.path("data", "mp02", glue("atus{file_base}_0324.dat"))

  if (!file.exists(file_name_out)) {
    zip_file <- file.path("data", "mp02", glue("atus{file_base}_0324.zip"))
    download.file(
      glue("https://www.bls.gov/tus/datafiles/atus{file_base}_0324.zip"),
      destfile = zip_file, quiet = TRUE, mode = "wb"
    )
    unzip(zip_file, exdir = file.path("data", "mp02"))
  }
  read_csv(file_name_out, show_col_types = FALSE)
}

df_resp      <- load_atus_data("resp")
df_rost      <- load_atus_data("rost")
df_act       <- load_atus_data("act")
df_act_codes <- read_excel(
  file.path("data", "mp02", "lexiconwex.xls"),
  sheet = "Lexicon", skip = 1
) |> select(activity_code = 1, activity_name = 2)
```

---

## Analysis

### 1. How Full-Time Workers Spend a Typical Workday

```{r}
#| label: worker-time
#| message: false
#| warning: false
#| fig-cap: "Average daily hours by activity category — full-time workers"

full_time_ids <- df_resp |>
  filter(TELFS == 1) |>
  pull(TUCASEID)

worker_activities <- df_act |>
  filter(TUCASEID %in% full_time_ids) |>
  left_join(df_act_codes, by = c("TRCODE" = "activity_code")) |>
  mutate(category = case_when(
    str_starts(as.character(TRCODE), "01") ~ "Work",
    str_starts(as.character(TRCODE), "03") ~ "Sleep",
    str_starts(as.character(TRCODE), "12") ~ "Leisure",
    str_starts(as.character(TRCODE), "07") ~ "Personal Care",
    TRUE ~ "Other"
  )) |>
  group_by(TUCASEID, category) |>
  summarize(minutes = sum(TUACTDUR24, na.rm = TRUE), .groups = "drop") |>
  group_by(category) |>
  summarize(avg_hours = mean(minutes / 60, na.rm = TRUE)) |>
  filter(category %in% c("Work", "Sleep", "Leisure", "Personal Care"))

worker_activities |>
  ggplot(aes(x = reorder(category, avg_hours), y = avg_hours, fill = category)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(avg_hours, 1)), hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c(
    Work = "#0d1b2a", Sleep = "#1e3a5f",
    Leisure = "#c9a84c", `Personal Care` = "#2e8b57"
  )) +
  scale_y_continuous(limits = c(0, 12), expand = c(0, 0)) +
  labs(
    title    = "How Full-Time Workers Spend a Typical Day",
    subtitle = "Average hours by activity category · ATUS 2003–2024",
    x = NULL, y = "Average Hours per Day"
  ) +
  theme_minimal(base_size = 13)
```

### 2. Work Time Trends Over 20 Years

```{r}
#| label: work-trend
#| fig-cap: "Average daily work hours for full-time workers, 2003–2024"

df_resp |>
  filter(TELFS == 1) |>
  select(TUCASEID, year = TUYEAR) |>
  left_join(
    df_act |>
      mutate(is_work = str_starts(as.character(TRCODE), "01")) |>
      group_by(TUCASEID) |>
      summarize(work_mins = sum(TUACTDUR24[is_work], na.rm = TRUE)),
    by = "TUCASEID"
  ) |>
  group_by(year) |>
  summarize(avg_work_hrs = mean(work_mins / 60, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg_work_hrs)) +
  geom_line(color = "#0d1b2a", linewidth = 1.2) +
  geom_point(color = "#c9a84c", size = 3) +
  scale_y_continuous(limits = c(6, 10)) +
  labs(
    title    = "Average Daily Work Hours — Full-Time Workers",
    subtitle = "ATUS 2003–2024 (excludes 2020 COVID year)",
    x = NULL, y = "Average Hours"
  ) +
  theme_minimal(base_size = 13)
```

### 3. Gender Differences in Leisure Time

```{r}
#| label: gender-leisure
#| fig-cap: "Average daily leisure hours by gender — full-time workers"

full_time_with_sex <- df_resp |>
  filter(TELFS == 1) |>
  select(TUCASEID, year = TUYEAR) |>
  left_join(df_rost |> filter(TULINENO == 1) |> select(TUCASEID, sex = TESEX),
            by = "TUCASEID")

full_time_with_sex |>
  left_join(
    df_act |>
      mutate(is_leisure = str_starts(as.character(TRCODE), "12")) |>
      group_by(TUCASEID) |>
      summarize(leisure_mins = sum(TUACTDUR24[is_leisure], na.rm = TRUE)),
    by = "TUCASEID"
  ) |>
  mutate(sex_label = if_else(sex == 1, "Men", "Women")) |>
  group_by(year, sex_label) |>
  summarize(avg_leisure = mean(leisure_mins / 60, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = avg_leisure, color = sex_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(Men = "#0d1b2a", Women = "#c9a84c")) +
  labs(
    title    = "Daily Leisure Hours by Gender — Full-Time Workers",
    subtitle = "ATUS 2003–2024",
    x = NULL, y = "Average Hours per Day", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### 4. The Ideal Day vs. Reality

This visualization compares what a balanced day might look like against what the
data shows full-time workers actually experience.

```{r}
#| label: activity-rings
#| fig-cap: "Target vs. Typical allocation of daily hours for a full-time worker"

worker_time <- worker_activities |>
  filter(category %in% c("Work", "Sleep", "Leisure")) |>
  rename(hours = avg_hours)

target <- tibble(
  category = c("Work", "Sleep", "Leisure"),
  hours    = c(8, 8, 4),
  type     = "Target (Ideal)"
)

typical <- worker_time |>
  mutate(type = "Typical (ATUS Data)")

rings <- bind_rows(target, typical) |>
  mutate(percent = hours / 24, remainder = 1 - percent) |>
  pivot_longer(c(percent, remainder), names_to = "part", values_to = "value")

ggplot(rings, aes(x = type, y = value, fill = interaction(category, part))) +
  geom_col(width = 0.85, color = NA) +
  coord_polar(theta = "y") +
  facet_wrap(~type) +
  scale_fill_manual(values = c(
    "Work.percent"     = "#0d1b2a", "Work.remainder"    = "#e8e8e8",
    "Sleep.percent"    = "#1e3a5f", "Sleep.remainder"   = "#e8e8e8",
    "Leisure.percent"  = "#c9a84c", "Leisure.remainder" = "#e8e8e8"
  )) +
  theme_void(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text       = element_text(size = 13, face = "bold"),
    plot.title       = element_text(size = 15, face = "bold", hjust = 0.5)
  ) +
  labs(title = "Ideal vs. Typical Day — Full-Time Workers", fill = "Activity")
```

---

## Key Takeaways

::: {.callout-tip title="What 20 years of data reveals"}
Full-time workers consistently work **more than 8 hours per day** when all work-related
time is included, leaving less room for sleep and leisure than most time-management
frameworks assume.

**Gender gaps in leisure persist**: men average roughly 30–45 more minutes of leisure
per day than women across the study period — a gap that has narrowed but not closed.

The COVID-19 period (2020) produced a sharp but temporary shift in time allocation,
particularly in commuting and in-person work, with partial reversion by 2022.
:::

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](index.html#projects)*
