---
title: "US Internal Migration — Forecasting Congressional Reapportionment"
page-layout: full
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
execute:
  eval: false
engine: knitr
---

::: {.proj-hero}
::: {.eyebrow}
Project 05 · R · Demographic Forecasting · Policy Analytics
:::

# US Internal Migration
### Forecasting 2032 Congressional Reapportionment

::: {.lede}
Every decade, the U.S. Census reshuffles congressional power based on where people live.
Using annual ACS migration flow data, this analysis builds a demographic model to
forecast 2030 state populations and predicts which states will gain or lose seats
in the next reapportionment.
:::

::: {.tag-row}
<span>R</span><span>tidycensus</span><span>ACS</span><span>Forecasting</span><span>Policy Analytics</span>
:::
:::

---

## Executive Summary

Using **ACS state-to-state migration flow data** combined with baseline population
estimates via tidycensus, this analysis forecasts 2030 state populations and applies
the **Huntington-Hill method** to apportion 435 House seats.

Key findings:

- Sun Belt states continue to show strong net in-migration
- Rust Belt and Northeast states face structural population outflows
- Small migration shifts can change apportionment for borderline states

---

## Data Acquisition

### State Population via tidycensus

```{r}
#| label: pop-data
#| eval: false
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(readxl)
library(gt)

data_dir <- file.path("data", "mp03")
if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

years <- c(2015:2019, 2021:2024)

state_pop <- map_dfr(years, function(yr) {
  get_acs(
    variables   = c(total_population = "B01003_001"),
    geography   = "state",
    year        = yr,
    survey      = "acs1",
    geometry    = FALSE,
    cache_table = TRUE
  ) |>
    mutate(year = yr) |>
    select(state_name = NAME, total_population = estimate, year)
})
```

### State-to-State Migration Flows

```{r}
#| label: migration-data
#| eval: false

parse_migration_table <- function(file_path, year_current, year_1y) {
  raw         <- read_excel(file_path, sheet = "Table", col_names = FALSE)
  header_row  <- as.character(raw[7, ])
  state_cols  <- seq(10, ncol(raw), by = 2)
  state_names <- header_row[state_cols]
  state_names <- state_names[state_names %in% state.name]
  state_cols  <- state_cols[header_row[state_cols] %in% state.name]

  data <- read_excel(file_path, sheet = "Table", skip = 8, col_names = FALSE) |>
    filter(!is.na(...1),
           !str_starts(...1, "United States"),
           !str_starts(...1, "Puerto Rico"))

  stationary <- data |>
    select(state_current = 1, population = 4) |>
    mutate(
      state_current = str_remove(state_current, "\\d+$") |> str_trim(),
      state_1y      = state_current,
      population    = as.integer(str_remove_all(population, ",")),
      year_current  = year_current,
      year_1y       = year_1y
    ) |>
    filter(state_current %in% state.name) |>
    select(state_current, state_1y, population, year_current, year_1y)

  flows <- map2_dfr(state_cols, state_names, function(col_idx, origin) {
    data |>
      select(state_current = 1, population = all_of(col_idx)) |>
      mutate(
        state_current = str_remove(state_current, "\\d+$") |> str_trim(),
        state_1y      = origin,
        population    = case_when(
          population == "N" ~ 0L,
          population == "X" ~ NA_integer_,
          TRUE ~ as.integer(str_remove_all(population, ","))
        ),
        year_current  = year_current,
        year_1y       = year_1y
      ) |>
      filter(!is.na(population), state_current %in% state.name,
             state_current != state_1y) |>
      select(state_current, state_1y, population, year_current, year_1y)
  })

  bind_rows(stationary, flows)
}

get_flow_file <- function(url, filename, year_current, year_1y) {
  fp <- file.path(data_dir, filename)
  if (!file.exists(fp)) download.file(url, destfile = fp, mode = "wb", quiet = TRUE)
  parse_migration_table(fp, year_current, year_1y)
}

flows_2024 <- get_flow_file(
  paste0("https://www2.census.gov/programs-surveys/demo/tables/",
         "geographic-mobility/2024/state-to-state-migration/",
         "State_to_State_Migration_Table_2024_T13.xlsx"),
  "state-to-state-2024.xlsx", 2024L, 2023L
)

flows_2023 <- get_flow_file(
  paste0("https://www2.census.gov/programs-surveys/demo/tables/",
         "geographic-mobility/2023/state-to-state-migration/",
         "State_to_State_Migration_Table_2023.xlsx"),
  "state-to-state-2023.xlsx", 2023L, 2022L
)

flows_all <- bind_rows(flows_2024, flows_2023)
```

---

## Analysis

### Net Migration by State

```{r}
#| label: net-migration
#| eval: false
#| fig-cap: "Top 15 states by net domestic in-migration 2023-2024"

net_migration <- flows_all |>
  filter(state_current != state_1y, !is.na(population)) |>
  group_by(state_current) |>
  summarize(
    in_flow  = sum(population[state_1y != state_current], na.rm = TRUE),
    out_flow = sum(population[state_1y == state_current], na.rm = TRUE),
    net      = in_flow - out_flow
  )

net_migration |>
  slice_max(abs(net), n = 15) |>
  mutate(direction = if_else(net > 0, "Net Gain", "Net Loss")) |>
  ggplot(aes(x = reorder(state_current, net), y = net / 1000, fill = direction)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Net Gain" = "#c9a84c", "Net Loss" = "#0d1b2a")) +
  scale_y_continuous(labels = scales::comma_format(suffix = "K")) +
  labs(title = "Top 15 States by Net Domestic Migration",
       subtitle = "Average of 2023 and 2024 ACS migration flows",
       x = NULL, y = "Net Migration (thousands)", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Congressional Apportionment — Huntington-Hill Method

```{r}
#| label: apportionment
#| eval: false

huntington_hill <- function(populations, n_seats = 435) {
  seats <- rep(1L, length(populations))
  names(seats) <- names(populations)
  for (i in seq_len(n_seats - length(populations))) {
    priority <- populations / sqrt(seats * (seats + 1))
    seats[which.max(priority)] <- seats[which.max(priority)] + 1L
  }
  seats
}

latest_pop <- state_pop |>
  filter(year == max(year)) |>
  select(state_name, pop_base = total_population)

forecast_2030 <- latest_pop |>
  left_join(net_migration |> rename(state_name = state_current), by = "state_name") |>
  mutate(
    net      = replace_na(net, 0),
    pop_2030 = pop_base + net * 6
  )

seats_now  <- huntington_hill(setNames(forecast_2030$pop_base,  forecast_2030$state_name))
seats_2030 <- huntington_hill(setNames(forecast_2030$pop_2030, forecast_2030$state_name))

tibble(
  state      = names(seats_2030),
  seats_now  = as.integer(seats_now),
  seats_2030 = as.integer(seats_2030),
  change     = as.integer(seats_2030) - as.integer(seats_now)
) |>
  filter(change != 0) |>
  arrange(desc(change)) |>
  gt() |>
  cols_label(state = "State", seats_now = "Current Seats",
             seats_2030 = "Projected 2030", change = "Change") |>
  tab_header(title = "Projected Congressional Seat Changes — 2032 Reapportionment") |>
  tab_style(style = cell_fill(color = "#eaf4ef"), locations = cells_body(rows = change > 0)) |>
  tab_style(style = cell_fill(color = "#fde8e8"), locations = cells_body(rows = change < 0))
```

---

## Key Takeaways

::: {.callout-tip title="What migration data tells us about political power"}
Congressional representation follows population — and population follows opportunity.
States gaining seats have **lower costs of living, warmer climates, and growing
job markets**. States at risk of losing seats can use data-driven recruitment
campaigns targeting high-outflow metros in neighboring states.
:::

---

*Analysis by Javier Bravo Bernal · [Back to Portfolio](index.html#projects)*
